Mutation loads in rheumatoid arthritis data from SComatic

### 2024-01-26 scomatic/reports/driver_coverage.Rmd
fix infercnv checkifexists issue, update driver coverage report, add knitting src

# iterate over each bam file and pileup specific regions
(
  echo -e 'chr\tpos\tdepth\tgene\tsample' ;
  for bam in $(find $input_dir -iname 'possorted_genome_bam.bam'); do
  	sample=$(echo $bam | cut -d "/" -f 10)
    while read -r gene coords ; do
      samtools depth -a -r $coords -Q 30 $bam |
      awk -v sample=$sample -v gene=$gene '{print $0"\t"gene"\t"sample}'
    done < <(sed 1d $drivers) ;
  done
) | cat > $output_dir/drivers_samtools_depth.tsv

@@

  # annotate exons
  exons <-
    ref_cds[c(drivers$gene)] %>%
    purrr::map(~ .x$intervals_cds %>% 
                 tibble::as_tibble() %>%
                 dplyr::transmute(chr = .x$chr, start = V1, end = V2) %>%
                 dplyr::mutate(exon = paste0('exon', dplyr::row_number()))) %>%
    dplyr::bind_rows(.id = 'gene') %>%
    dplyr::mutate(pos = purrr::map2(start, end, ~ .x:.y)) %>%
    tidyr::unnest(cols = 'pos') %>%
    dplyr::select(-start, -end)
  cov <-
    cov %>%
    dplyr::left_join(exons) %>%
    dplyr::mutate(exon = ifelse(is.na(exon), 'intron', exon),
                  feature = ifelse(exon == 'intron', 'intron', 'exon'))

@@

# function: plot all
plot_all <- function(p_df, p_exon_coords, info) {
  ggplot() +
    geom_rect(
      data = p_exon_coords,
      aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
      colour = 'lightblue', fill = 'lightblue'
    ) +
    geom_col(
      data = p_df,
      aes(x = pos, y = depth, fill = feature)) +
    theme_classic() +
    scale_fill_manual(values = c(exon = 'darkblue', intron = 'darkgrey')) +
    labs(title = paste(info, '(all)'))
}

# function: plot exons 
plot_exons <- function(p_df, title) {
  p_df %>%
    # get only exonic sequence
    dplyr::filter(feature == 'exon') %>%
    # fix exon ordering
    dplyr::mutate(
      exon = exon %>% forcats::fct_reorder(exon %>% gsub('exon', '', .) %>% as.numeric())
    ) %>%
    ggplot(aes(x = pos, y = depth)) +
    geom_col(fill = 'darkblue') +
    facet_grid(. ~ exon, space = 'free_x', scales = 'free_x') +
    theme_classic() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.spacing = unit(.05, "lines"),
          panel.border = element_rect(fill = NA)) +
    labs(title = paste(info, '(exons)'))
}

# create list of plots by gene x sample
p <- xfun::cache_rds({
  # initiate plot list
  p <- list()

@@

      list(by_sample = list(lvl = 'drivers_samtools_depth', title_suffix = ''),
           by_celltype = list(lvl = 'drivers_samtools_depth', title_suffix = '_by_celltype')) %>%
        purrr::map(function(x){

          # plot all
          p_all <-
            plot_all(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              p_exon_coords = exon_coords %>% dplyr::filter(gene == i$gene),
              paste0(info, x$title_suffix)
            )

          # plot exons only
          p_all <<-
            plot_exons(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              paste0(info, x$title_suffix)
            )

          # facet grid for celltypes
          if (title_suffix == '_by_celltype') {
            p_all <- p_all + facet_grid(~ celltype)
            p_exons <- p_exons + facet_grid(~ celltype)
          }

          # assign globally
          p[[info]][[paste0('all', title_suffix)]] <<- p_all
          p[[info]][[paste0('exons', title_suffix)]] <<- p_exons

@@

for (info in names(p)) {
  chunks <- c(chunks, paste0('\n####', info))

  knit_plot <- c(
    '',
    paste0('```{r ', info, ', class.source = "fold-hide", warning = F, message = F, ',
           'fig.dim = c(20, ', 3 * length(p[[info]]), ')}'),
    paste0('do.call("grid.arrange", c(p[["', info, '"]], ncol = 1))')
    '```',
    '')
  knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
  dev_off_if()

@@

>>>>

# iterate over each bam file and pileup specific regions
(
  echo -e 'chr\tpos\tdepth\tgene\tsample' ;
  for bam in $(find $input_dir -iname 'possorted_genome_bam.bam'); do
  	sample=$(echo $bam | cut -d "/" -f 10 | sed 's/\-/_/g')
    while read -r gene coords ; do
      samtools depth -a -r $coords -Q 30 $bam |
      awk -v sample=$sample -v gene=$gene '{print $0"\t"gene"\t"sample}'
    done < <(sed 1d $drivers) ;
  done
) | cat > $output_dir/drivers_samtools_depth.tsv

@@

  # annotate exons
  exons <-
    ref_cds[c(drivers$gene)] %>%
    purrr::map(~ .x$intervals_cds %>% 
                 tibble::as_tibble() %>%
                 dplyr::transmute(chr = .x$chr, start = V1, end = V2) %>%
                 dplyr::mutate(exon = paste0('exon', dplyr::row_number()))) %>%
    dplyr::bind_rows(.id = 'gene') %>%
    dplyr::mutate(pos = purrr::map2(start, end, ~ .x:.y)) %>%
    tidyr::unnest(cols = 'pos') %>%
    dplyr::select(-start, -end)
  cov <-
    cov %>%
    dplyr::left_join(exons, by = c('chr', 'pos', 'gene')) %>%
    dplyr::mutate(exon = ifelse(is.na(exon), 'intron', exon),
                  feature = ifelse(exon == 'intron', 'intron', 'exon'))

@@

# function: plot all
plot_all <- function(p_df, p_exon_coords, p_title) {
  ggplot() +
    geom_rect(
      data = p_exon_coords,
      aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
      colour = 'lightblue', fill = 'lightblue'
    ) +
    geom_col(
      data = p_df,
      aes(x = pos, y = depth, fill = feature)) +
    theme_classic() +
    theme(panel.border = element_rect(fill = NA),
          panel.spacing = unit(.05, 'lines')) +
    scale_fill_manual(values = c(exon = 'darkblue', intron = 'darkgrey')) +
    labs(title = paste(p_title, '(all)')) +
    facet_grid(celltype ~ ., space = 'free_x', scales = 'free_x') 
}

# function: plot exons 
plot_exons <- function(p_df, p_title) {
  p_df %>%
    # get only exonic sequence
    dplyr::filter(feature == 'exon') %>%
    # fix exon ordering
    dplyr::mutate(
      exon = exon %>% forcats::fct_reorder(exon %>% gsub('exon', '', .) %>% as.numeric())
    ) %>%
    ggplot(aes(x = pos, y = depth, group = celltype)) +
    geom_col(fill = 'darkblue') +
    theme_classic() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.spacing = unit(.05, 'lines'),
          panel.border = element_rect(fill = NA)) +
    labs(title = paste(p_title, '(exons)')) +
    facet_grid(celltype ~ exon, space = 'free_x', scales = 'free_x') 
}

# create list of plots by gene x sample
p <- xfun::cache_rds({

  # initiate plot list
  p <- list()
  n_cts <- list()

@@

      # get number of celltypes for plot dims
      p[[info]]$n_cts <<- 
        df_i %>% 
        dplyr::filter(!is.na(celltype)) %>% 
        dplyr::pull(celltype) %>% 
        dplyr::n_distinct()

      # plot by sample and by celltype
      list(by_sample = list(lvl = 'drivers_samtools_depth', title_suffix = ''),
           by_celltype = list(lvl = 'drivers_samtools_depth_by_celltype', title_suffix = '_by_celltype')) %>%
        purrr::map(function(x){

          # plot all
          p[[info]][[paste0('all', x$title_suffix)]] <<-
            plot_all(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              p_exon_coords = exon_coords %>% dplyr::filter(gene == i$gene),
              p_title = paste0(info, x$title_suffix)
            )

          # plot exons only
          p[[info]][[paste0('exons', x$title_suffix)]] <<-
            plot_exons(
              p_df = df_i %>% dplyr::filter(lvl == x$lvl),
              p_title = paste0(info, x$title_suffix)
            )

@@

for (info in names(p)) {
  chunks <- c(chunks, paste0('\n####', info))
  message(info)

  knit_plot <- c(
    '',
    paste0('```{r ', info, ', class.source = "fold-hide", warning = F, message = F, ',
           'fig.dim = c(20, ', 4 + 4 * p[[info]]$n_cts, ')}'),
    paste0('p_i <- p[["', info, '"]]'),
    paste0('grid.arrange(p_i$all, p_i$all_by_celltype, p_i$exons, p_i$exons_by_celltype, ncol = 1, ',
           'heights = c(2, p_i$n_cts * 2, 2, p_i$n_cts * 2))'),
    '```',
    '')
  knitted_plot <- knitr::knit_child(text = knit_plot, envir = environment(), quiet = T)
  dev_off_if()

@@

```{r mean_coverage}
cov %>%
  dplyr::filter(lvl == 'drivers_samtools_depth') %>%
  dplyr::group_by(sample, gene, feature) %>%
  dplyr::summarise(mean_depth = mean(depth)) %>%
  ggplot(aes(x = sample, y = mean_depth, fill = feature)) +
  geom_col(position = 'dodge') +
  facet_grid(gene ~ .) +
  theme_classic()
```
